---
title: "Analyzing Temporal Factors Influencing Fish Concentration - Using R and Python"
author: "Yingwen"
date: "2026-09-21"
---

## Data Source

The data used in this post comes from the `fishing` dataset in the [`COUNT` package on CRAN](https://cran.r-project.org/package=COUNT). It contains data detailing fish catching data such as site number, fish density and so on.

```{python}
#| label: setup_python
#| message: false
#| warning: false

import pandas as pd

# Load dataset from CRAN source repository
url = "https://raw.githubusercontent.com/vincentarelbundock/Rdatasets/master/csv/COUNT/fishing.csv"
fishing = pd.read_csv(url)

# Display initial rows
fishing.head()
```

```{r}
#| label: setup_r
#| message: false
#| warning: false
library(reticulate)
library(ggplot2)
```

## Exploratory Analysis

To understand the temporal factors influencing fish concentration, we first examine how annual observations impact average fish density.

```{python}
#| label: year-summary-stats
#| echo: true

year_summary = (
    fishing.groupby("year")
    .agg(
        count=("density", "size"),
        mean_density=("density", "mean"),
        median_density=("density", "median")
    )
    .sort_values(by="mean_density", ascending=False)
    .reset_index()
)

year_summary
```
Evaluating aggregate annual density yields no clear directional trend. Next, we aggregate density by individual observation period to assess intra-annual patterns.

```{python}
#| label: period-summary-stats
#| echo: true

period_summary = (
    fishing.groupby("period")
    .agg(
        count=("density", "size"),
        mean_density=("density", "mean"),
        median_density=("density", "median")
    )
    .sort_values(by="mean_density", ascending=False)
    .reset_index()
)

period_summary
```
Grouping by period reveals a pronounced downward trajectory in fish density over time. To better evaluate why annual averages mask this trend, we generate a figure comparing yearly density values.

```{r}
# Convert year to factor so ggplot treats it as categorical discrete groups on the x-axis
ggplot(py$fishing, aes(x = factor(year), y = density)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  labs(
    title = "Yearly Fish Density Distribution",
    x = "Year",
    y = "Fish Density"
  ) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

## What We Found

Analyzing `@fig-yearly-density` shows key insights into the temporal pattern of fish density across the dataset:

- **Inconsistent Sampling Volume:** Data collection is heavily concentrated in 1978 to 1983 and 2000 to 2002. Years such as 1977, 1984, 1986, and 1989 contain very few observations, resulting in near-zero spread.
- **Higher Variability in early 1980s:** Between 1979 and 1983, fish density exhibited high overall variance, with median values near 0.005 and frequent high-density peaks exceeding 0.02.