---
title: "Analyzing Choice of Fishing Modes"
author: "Yingwen"
date: "2026-09-21"
---

## Data Source

The data used in this post comes from the `fishing` dataset in the [`COUNT` package on CRAN](https://cran.r-project.org/package=COUNT). It contains survey data detailing recreational fishing mode choices (beach, pier, boat, or charter) alongside individuals' income and target species catch rates.

```{r}
#| label: setup
#| message: false
#| warning: false

library(COUNT)
library(dplyr)
library(ggplot2)

# Load dataset into memory
data("fishing", package = "COUNT")
head(fishing)
```

## Exploratory Analysis

To understand the temporal factors influencing fish concentration, we first examine how annual observations impact average fish density.

```{r}
#| label: year-summary-stats
#| echo: true

mode_summary <- fishing |>
  group_by(year) |>
  summarize(
    count = n(),
    mean_density = mean(density, na.rm = TRUE),
    median_density = median(density, na.rm = TRUE)
  ) |>
  arrange(desc(mean_density))

mode_summary
```
Evaluating aggregate annual density yields no clear directional trend. Next, we aggregate density by individual observation period to assess intra-annual patterns.

```{r}
#| label: period-summary-stats
#| echo: true

mode_summary <- fishing |>
  group_by(period) |>
  summarize(
    count = n(),
    mean_density = mean(density, na.rm = TRUE),
    median_density = median(density, na.rm = TRUE)
  ) |>
  arrange(desc(mean_density))

mode_summary
```
Grouping by period reveals a pronounced downward trajectory in fish density over time. To better evaluate why annual averages mask this trend, we generate a figure comparing yearly density values.

```{r}
#| label: fig-yearly-density
#| fig-cap: "Distribution of fish density across observation years."
#| echo: true

ggplot(fishing, aes(x = factor(year), y = density)) +
  geom_boxplot(fill = "steelblue", alpha = 0.4, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, color = "darkblue") +
  labs(
    x = "Year",
    y = "Fish Density",
    title = "Yearly Fish Density Distribution"
  ) +
  theme_minimal()
```

## What We Found

Analyzing `@fig-yearly-density` shows key insights into the temporal pattern of fish density across the dataset:

- **Inconsistent Sampling Volume:** Data collection is heavily concentrated in 1978 to 1983 and 2000 to 2002. Years such as 1977, 1984, 1986, and 1989 contain very few observations, resulting in near-zero spread.
- **Higher Variability in early 1980s:** Between 1979 and 1983, fish density exhibited high overall variance, with median values near 0.005 and frequent high-density peaks exceeding 0.02.