---
title: "Analyzing Relationship between Temperature and Relative Humidity across NYC Airports"
author: "Yingwen"
date: "2026-09-22"
---

## Data Source

The data used in this post comes from the `weather` dataset in the [`nycflights13` package on CRAN](https://cran.r-project.org/package=nycflights13). It contains the weather information across airports near New York.

```{r}
#| label: setup
#| message: false
#| warning: false

library(nycflights13)
library(dplyr)
library(ggplot2)

# Load dataset into memory
data("weather", package = "nycflights13")
head(weather)
```

## Exploratory Analysis

To evaluate temperature differences across the metropolitan region, we filter missing readings and calculate central tendency and spread for each airport.

```{r}
#| label: summary-stats
#| echo: true

# Clean missing values and compute summary metrics
weather |>
  filter(!is.na(temp)) |>
  group_by(origin) |>
  summarise(
    Mean_Temp = round(mean(temp), 2),
    Median_Temp = round(median(temp), 2),
    SD_Temp = round(sd(temp), 2),
    IQR_Temp = round(IQR(temp), 2),
    Observations = n()
  )
```

We generate comparative boxplots overlaid with density contours to visualize the spread, skewness, and potential outliers across the three airports.

```{r}
#| label: fig-temp-distributions
#| fig-cap: "Temperature distributions and medians across EWR, JFK, and LGA airports."
#| echo: true

weather_clean <- weather |> filter(!is.na(temp))

ggplot(weather_clean, aes(x = origin, y = temp, fill = origin)) +
  geom_boxplot() +
  labs(
    title = "Temperature Distribution by NYC Airport",
    subtitle = "Hourly records from 2013",
    x = "Airport Code",
    y = "Temperature (°F)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Although the means appear similar, we run an ANOVA to formally test whether the minor observed differences in mean temperature across airports are statistically significant.

```{r}
#| label: temp-anova-test
#| echo: true

# Fit ANOVA model testing temperature differences across origins
anova_model <- aov(temp ~ origin, data = weather_clean)
summary(anova_model)
```
The ANOVA test yields an $F$-statistic of 13.16 with a $p$-value of 1.95e-06. Because the $p$-value is very small, we reject the null hypothesis. There is statistically significant difference in mean air temperature among EWR, JFK, and LGA.